Combinatorial interpretations of holonomic recurrences

### 2021-09-05 article/main.tex
Add fixups

\RequirePackage{geometry}
\RequirePackage{tabularx}
\RequirePackage{longtable}
\RequirePackage{graphicx}
\RequirePackage{svg}

@@

\begin{figure}[h]
    \begin{center}
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B\)\\
                \includesvg[scale=0.5]{binary__def_1}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(=\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(\text{leaf}\)\\
                \includesvg[scale=0.5]{binary__def_2}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(+\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B^2\)\\
                \includesvg[scale=0.5]{binary__def_3}%
            \end{center}
        \end{minipage}%
    \end{center}
    \caption{Binary tree recursive definition}
    \label{fig:binary_recursion}
\end{figure}

@@

\begin{example}
    \label{ex:bin_gf}
    Having our recursive definition of the object, we can easily construct a generating function for it. Let's create a generating function that will count the number of full binary trees consisting of exactly \(k\) internal nodes.\todo{define sequence for the structure}

Note that we used \(z\) to mark internal nodes. Leaves are of size \(0\), so they are marked using \(z^0 = 1\).

@@

\[\begin{array}{rcl}
        \gf{B}(z) &=& 1 + z\gf{B}^2(z)\\
                  &=& 1 + z\bigg(1 + z\gf{B}^2(z)\bigg)\bigg(1 + z\gf{B}^2(z)\bigg)\\
                  &=& 1 + \mhl{z}\bigg(1 + \mhl{z}\Big(\mhl{1} + z\gf{B}^2(z)\Big)\Big(\mhl{1} + z\gf{B}^2(z)\Big)\bigg)\bigg(\mhl{1} + z\gf{B}^2(z)\bigg)
\end{array}\]

The highlighted terms describe following tree:

\begin{center}
    \includesvg[scale=.7]{intro__iilll}
\end{center}
\end{example}

@@

Most of the time we will be working with holonomic recurrences, let's introduce them.

\begin{definition}[\cite{holotoolkit}]
    \label{def:holo_1}
    An infinite sequence \(A = (a_i)_{i=0}^{\infty}\) is holonomic (aka. P-finite, P-recursive or D-recursive) \(\textiff\) there exists a holonomic recurrence for it, i.e.:

\[\exists_{r \in \N} \exists_{p_0, \ldots, p_r \in \poly} \forall_{n \in \N} \sum_{i=0}^r p_i(n)a_{n+i} = 0\]
\end{definition}

@@

\begin{example}
    The sequence \(a_n = \frac{5n - 3}{3n + 5}\) is holonomic, because it satisifies:
    \[(3n + 5)(5n + 2) a_n - (5n - 3)(3n + 8) a_{n+1} = 0\]
\end{example}

@@

The equivalence of these two definitions assures us that if we find generating function describing the desired objects satisfying definition\ \ref{def:holo_2}, we will be able to find holonomic recurrence mentioned in definition\ \ref{def:holo_1}~\cite{complexity}.

\begin{example}
    \label{ex:gfun-rec}
    For generating function full binary trees defined in example\ \ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN~\cite{gfun} to immediately find holonomic recurrence of our interest:

@@

\begin{figure}[h]
    \centering
    \includesvg[scale=.7]{intro__oononnn}
    \caption{\(\n \n \no \n \no \no \no\)}
    \label{fig:oononnn}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_left}\end{minipage}%
    \caption{\(\enc{R_l(T, v)}\)}
    \label{fig:remy_left}
\end{figure}

\begin{figure}[h]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_right}\end{minipage}%
    \caption{\(\enc{R_r(T, v)}\)}
    \label{fig:remy_right}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includeinlinesvg{.2}{unary_binary__base}%
    \(\Rightarrow\)%
    \includeinlinesvg{.3}{unary_binary__full}
    \caption{Unary-binary tree to full-binary tree transformation}
    \label{fig:unary_binary_transformation}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includesvg[scale=0.4]{schroder__example_1}
    \includesvg[scale=0.4]{schroder__example_2}
    \includesvg[scale=0.4]{schroder__example_3}
    \includesvg[scale=0.4]{schroder__example_4}
    \(\ldots\)
    \caption{Schröder tree node}
    \label{fig:schroder_trees}
\end{figure}

>>>>

\RequirePackage{geometry}
\RequirePackage{tabularx}
\RequirePackage{longtable}
\RequirePackage{graphicx}
\RequirePackage{svg}
\RequirePackage{float}

@@

\begin{figure}[H]
    \begin{center}
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B\)\\
                \includesvg[scale=0.5]{binary__def_1}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(=\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(\text{leaf}\)\\
                \includesvg[scale=0.5]{binary__def_2}%
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.05\textwidth}
            \begin{center}
                \(+\)\\
            \end{center}
        \end{minipage}%
        \begin{minipage}[t]{.2\textwidth}
            \begin{center}
                \(B^2\)\\
                \includesvg[scale=0.5]{binary__def_3}%
            \end{center}
        \end{minipage}%
    \end{center}
    \caption{Binary tree recursive definition}
    \label{fig:binary_recursion}
\end{figure}

@@

If we have sequence counting some objects and corresponding generating function, we say that this function counts the objects.

\begin{example}
    \label{ex:bin_gf}
    Let's consider recursive definition of binary trees shown in Figure~\ref{fig:binary_recursion}. We can create a functional equation to count trees with exactly n internal nodes. To do so, we make internal nodes contribute to the generating function with \(z^1 = z\) and leaves with \(z^0 = 1\):

\[\gf{B}(z) = 1 + z\gf{B}^2(z).\]

@@

\[\begin{array}{rclr}
        \gf{B}(z) &=& 1 + z\gf{B}^2(z)\\
                  &=& 1 + z\bigg(1 + z\gf{B}^2(z)\bigg)\bigg(1 + z\gf{B}^2(z)\bigg)\\
                  &=& 1 + \mhl{z}\bigg(1 + \mhl{z}\Big(\mhl{1} + z\gf{B}^2(z)\Big)\Big(\mhl{1} + z\gf{B}^2(z)\Big)\bigg)\bigg(\mhl{1} + z\gf{B}^2(z)\bigg) &(*)\\
                  &=& 1 + z\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\bigg(1 + z\Big(1 + z\gf{B}^2(z)\Big)\Big(1 + z\gf{B}^2(z)\Big)\bigg)\\
                  &=& 1 + z + 2 z^2 + \bigg(2 \gf{B}^2(z) + \Big(1 + z \gf{B}^2(z)\Big)^2\bigg) z^3
\end{array}\]

It is enough, as the term \(\bigg(2 \gf{B}^2(z) + \Big(1 + z \gf{B}^2(z)\Big)^2\bigg) z^3\) will contribute only to \(z^k\) where \(k > 2\). Therefore, we can see that there are \(2\) trees with exactly \(2\) internal nodes, as multiplier next to \(z^2\) is \(2\).

To visualize even more, let's take a look at the line marked with \((*)\). The highlighted terms describe following tree:

\begin{figure}[H]
    \begin{center}
        \includesvg[scale=.7]{intro__iilll}
    \end{center}
    \caption{Tree corresponding to highlighted terms in \((*)\)}
    \label{fig:tree_corresponding_to_mhl}
\end{figure}

@@

Since most of the time we will be working with holonomic recurrences, let's introduce them.

\begin{definition}[\cite{holotoolkit}]
    \label{def:holo_1}
    An infinite sequence \(A = (a_i)_{i=0}^{\infty}\) is holonomic (aka. P-finite, P-recursive or D-recursive) \(\textiff\) there exists a holonomic recurrence for it, i.e.

\[\exists_{r \in \N} \exists_{p_0, \ldots, p_r \in \poly} \forall_{n \in \N} \sum_{i=0}^r p_i(n)a_{n+i} = 0 .\]
\end{definition}

@@

\begin{example}
    The sequence \(a_n = \frac{5n - 3}{3n + 5}\) is holonomic, because it satisifies
    \[(3n + 5)(5n + 2) a_n - (5n - 3)(3n + 8) a_{n+1} = 0 .\]
\end{example}

@@

\todo{Continue}
The equivalence of these two definitions assures us that if we find generating function describing the desired objects satisfying definition\ \ref{def:holo_2}, we will be able to find holonomic recurrence mentioned in definition\ \ref{def:holo_1}~\cite{complexity}.

\begin{example}
    \label{ex:gfun-rec}
    For generating function full binary trees defined in Example~\ref{ex:bin_gf}, i.e.:
    \[\gf{B}(z) = 1 + z\gf{B}^2(z)\]
    we can use GFUN~\cite{gfun} to immediately find holonomic recurrence of our interest:

@@

\begin{figure}[H]
    \centering
    \includesvg[scale=.7]{intro__oononnn}
    \caption{\(\n \n \no \n \no \no \no\)}
    \label{fig:oononnn}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_left}\end{minipage}%
    \caption{\(\enc{R_l(T, v)}\)}
    \label{fig:remy_left}
\end{figure}

\begin{figure}[H]
    \centering
    \begin{minipage}{.25\textwidth}\includesvg[scale=0.35]{binary__remy_base}\end{minipage}%
    \(\Rightarrow\)
    \begin{minipage}{.4\textwidth}\includesvg[scale=0.5]{binary__remy_right}\end{minipage}%
    \caption{\(\enc{R_r(T, v)}\)}
    \label{fig:remy_right}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includeinlinesvg{.2}{unary_binary__base}%
    \(\Rightarrow\)%
    \includeinlinesvg{.3}{unary_binary__full}
    \caption{Unary-binary tree to full-binary tree transformation}
    \label{fig:unary_binary_transformation}
\end{figure}

@@

\begin{figure}[H]
    \centering
    \includesvg[scale=0.4]{schroder__example_1}
    \includesvg[scale=0.4]{schroder__example_2}
    \includesvg[scale=0.4]{schroder__example_3}
    \includesvg[scale=0.4]{schroder__example_4}
    \(\ldots\)
    \caption{Schröder tree node}
    \label{fig:schroder_trees}
\end{figure}
